---
title: "EDA on Admissions Data"
date: last-modified
author: "Courtney St. Onge & Seira Ramchandani"
format: html
jupyter: python3
---

```{python}
#import libraries
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
```

```{python}
# Importing the data

df = pd.read_csv("../../data/processed/capstone_data_processed.csv")
```

```{r}
library(tidyverse)
df = read_csv("C:\\Users\\cocom\\Downloads\\capstone_data_processed.csv")
```

## Data Exploration
This is a print-out of the columns, what they are, and what they represent. 

The columns that represent our target variables are 
- is_nw5
- is_large_public
- dest_other

These three columns are all booleans, with a 1 representing a truth and a zero representing false. Each row should have 1 out of the three columns that have a 1, and the other two columns should have a zero, since a school can either be a NW5 school, a large public school, or another type of school. 

This cell confirms that there are no instances in which a person attends a school that is marked as multiple of the school types. 
```{python}
df.loc[((df['is_nw5']==1) & (df['is_large_public']==1)) | ((df['is_nw5']==1) & (df['dest_other'] == 1)) | ((df['is_large_public']==1) & df['dest_other']==1) | (df['dest_other']==1)& (df['is_nw5']==1) & (df['is_large_public']==1)]
```


```{python}
df.head()
#df.names()
```

## Distribution of Variables

This section will look at the distribution of variables faceted by NW5 school, large public school, or another type of school.


```{python}
g=sns.pairplot(df, diag_kind='kde')
plt.show()
```

### Categorical Variables


#### Geography

Within each student's application, we have access to information about where they lived when they applied to Willamette.

```{r}
# graph of where students are from by region
place_df = df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        ),
        n_region = case_when(
            active_region %in% c('CA', 'WA', 'OR') ~ "West Coast",
            active_region %in% c('AZ', 'NV', 'NM', 'TX' ) ~ 'Southwest',
            active_region %in% c('ID', 'MT', 'UT', 'WY', 'CO') ~ 'Rocky Mtn',
            active_region %in% c('ND', 'SD', 'NE', 'KS', 'OK', 'MN', 'IA', 'MO', 'IL', 'OH', 'MI', 'WI', 'IN' ) ~ 'Midwest',
            active_region %in% c('MA', 'VT', 'MA', 'NH', 'NY', 'RI', 'NJ', 'CT', 'DE', 'MD', 'PA', 'DC') ~ 'Northeast',
            active_region %in% c('AR', 'AL', 'LA', 'MS', 'GA', 'SC', 'NC', 'VA', 'WV', 'KY', 'TN', 'FL') ~ 'South',
            active_region %in% c("AK", 'HI') ~ 'AK/HI',
            (is.na(active_region) & active_country != 'United States' & !is.na(active_country)) | (active_country != 'United States' & !is.na(active_region)) ~ 'International') ) %>%
            select('school_loc', 'n_region', 'active_region', 'active_country') %>%
        drop_na(n_region)
  #print(place_df)
ggplot(data=place_df, aes(x=fct_rev(fct_infreq(n_region)))) + geom_bar() +coord_flip() +labs(
  title="Region of Origin for Applicants",
  y = 'Number of Applicants',
  x = 'Applicant Region of Origin'
)+ geom_text(stat = "count", aes(label = after_stat(count)), hjust = -.23)

```


```{r}
# split by region including other by count
place_df = df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        ),
        n_region = case_when(
            active_region %in% c('CA', 'WA', 'OR') ~ "West Coast",
            active_region %in% c('AZ', 'NV', 'NM', 'TX' ) ~ 'Southwest',
            active_region %in% c('ID', 'MT', 'UT', 'WY', 'CO') ~ 'Rocky Mtn',
            active_region %in% c('ND', 'SD', 'NE', 'KS', 'OK', 'MN', 'IA', 'MO', 'IL', 'OH', 'MI', 'WI', 'IN' ) ~ 'Midwest',
            active_region %in% c('MA', 'VT', 'MA', 'NH', 'NY', 'RI', 'NJ', 'CT', 'DE', 'MD', 'PA', 'DC') ~ 'Northeast',
            active_region %in% c('AR', 'AL', 'LA', 'MS', 'GA', 'SC', 'NC', 'VA', 'WV', 'KY', 'TN', 'FL') ~ 'South',
            active_region %in% c("AK", 'HI') ~ 'AK/HI',
            (is.na(active_region) & active_country != 'United States' & !is.na(active_country)) | (active_country != 'United States' & !is.na(active_region)) ~ 'International')
        )  %>% drop_na(n_region)
ggplot(data=place_df, aes(x=n_region, fill=school_loc)) + geom_bar(stat="count", position="dodge") 

```

```{r}
# graph of percentages of students split by region attending each type of school
df_loc = df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        ),
        n_region = case_when(
            active_region %in% c('CA', 'WA', 'OR') ~ "West Coast",
            active_region %in% c('AZ', 'NV', 'NM', 'TX' ) ~ 'Southwest',
            active_region %in% c('ID', 'MT', 'UT', 'WY', 'CO') ~ 'Rocky Mtn',
            active_region %in% c('ND', 'SD', 'NE', 'KS', 'OK', 'MN', 'IA', 'MO', 'IL', 'OH', 'MI', 'WI', 'IN' ) ~ 'Midwest',
            active_region %in% c('MA', 'VT', 'MA', 'NH', 'NY', 'RI', 'NJ', 'CT', 'DE', 'MD', 'PA', 'DC') ~ 'Northeast',
            active_region %in% c('AR', 'AL', 'LA', 'MS', 'GA', 'SC', 'NC', 'VA', 'WV', 'KY', 'TN', 'FL') ~ 'South',
            active_region %in% c("AK", 'HI') ~ 'AK/HI',
            (is.na(active_region) & active_country != 'United States' & !is.na(active_country)) | (active_country != 'United States' & !is.na(active_region)) ~ 'International')
        ) %>% drop_na(n_region, school_loc) %>% 
    group_by(n_region, school_loc) %>% 
    summarise(subtotal = n(), .groups = "drop_last") %>% 
    mutate(total_by_var1 = sum(subtotal),
          percent = subtotal/total_by_var1)
        
    print(df_loc)
    ggplot(data=df_loc,aes(x=n_region, y=percent,fill=school_loc)) + geom_col( position="dodge") +
    scale_y_continuous(labels = scales::percent)
```

This graph shows us that 

```{r}
# graph of percentages of students who go to either large public or nw5, not other. split by region
df_loc = df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        ),
        n_region = case_when(
            active_region %in% c('CA', 'WA', 'OR') ~ "West Coast",
            active_region %in% c('AZ', 'NV', 'NM', 'TX' ) ~ 'Southwest',
            active_region %in% c('ID', 'MT', 'UT', 'WY', 'CO') ~ 'Rocky Mtn',
            active_region %in% c('ND', 'SD', 'NE', 'KS', 'OK', 'MN', 'IA', 'MO', 'IL', 'OH', 'MI', 'WI', 'IN' ) ~ 'Midwest',
            active_region %in% c('MA', 'VT', 'MA', 'NH', 'NY', 'RI', 'NJ', 'CT', 'DE', 'MD', 'PA', 'DC') ~ 'Northeast',
            active_region %in% c('AR', 'AL', 'LA', 'MS', 'GA', 'SC', 'NC', 'VA', 'WV', 'KY', 'TN', 'FL') ~ 'South',
            active_region %in% c("AK", 'HI') ~ 'AK/HI',
            (is.na(active_region) & active_country != 'United States' & !is.na(active_country)) | (active_country != 'United States' & !is.na(active_region)) ~ 'International')
        ) %>% drop_na(n_region, school_loc) %>% 
    group_by(n_region, school_loc) %>% 
    summarise(subtotal = n(), .groups = "drop_last") %>% 
    mutate(total_by_var1 = sum(subtotal),
          percent = subtotal/total_by_var1) %>%
          #filtering out the "other schools"
          filter(school_loc == 'Large Public' | school_loc == 'NW 5') 
    print(df_loc)
    ggplot(data=df_loc,aes(x=n_region, y=percent,fill=school_loc)) + geom_col(position="dodge") +
    scale_y_continuous(labels = scales::percent) +
    labs (
      x= "Applicant Region of Origin",
      y = "Perentage of Students per Region",
      fill = "Attended School Category"
      )
```

From this graph, we can see that

#### Financials

The cost to attend college can be quite large, especially going to a NW5 school instead of a large public school. This has the potential to be a large factor in where students decide to attend. 
It could also be interesting to look at if students are first generation and if they tend to go to a certain type of school, potentially one with less overall cost to them and their families. 


Columns to look at:
common_app_fee_waiver - could be null or zero even if they would receive financial aid, but they applied to schools that did not have an application fee
need_based_aid - could be indicative of a person who would have to pay attention to total cost before they choose a school.


```{r}
df %>% filter('need_based_aid' != 0) %>% select('need_based_aid', 'talent_scholarships', 'efc_sai', 'common_app_fee_waiver')
```
Based on the columns that are null and active, we will look at the potential correlation between need based aid and destination, as well as common_app_fee_waiver and destination.

```{r}

df_loc = df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        )) %>% 
    drop_na(need_based_aid, school_loc) %>% 
    group_by(need_based_aid, school_loc) %>% 
    summarise(subtotal = n(), .groups = "drop_last") %>% 
    mutate(total_by_var1 = sum(subtotal),
          percent = subtotal/total_by_var1) %>%
    mutate(school_loc = fct_reorder(school_loc, percent, .fun = identity, .desc=TRUE))
          #filtering out the "other schools"
          #filter(school_loc == 'Large Public' | school_loc == 'NW 5') 
    print(df_loc)
    ggplot(data=df_loc,aes(x=as.factor(need_based_aid), y=percent,fill=school_loc)) + geom_col( position="dodge") +
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(labels = c("0" = "Not Eligible", "1" = "Eligible")) +
    labs (
      x= "Eligibility of Need Based Aid",
      y = "Percentage of Students",
      fill = "Attended School Category"
      )+ geom_text(aes(label = round(percent*100,2)), vjust = -.5, position = position_dodge(width=0.9))
```

```{r}
df_loc = df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        )) %>% 
    drop_na(need_based_aid, school_loc) %>% 
    group_by(need_based_aid, school_loc) %>% 
    summarise(subtotal = n(), .groups = "drop_last") %>% 
    mutate(total_by_var1 = sum(subtotal),
          percent = subtotal/total_by_var1) %>%
    mutate(school_loc = fct_reorder(school_loc, percent, .fun = identity, .desc=TRUE)) %>%
          #filtering out the "other schools"
          filter(school_loc == 'Large Public' | school_loc == 'NW 5') 
    print(df_loc)
    ggplot(data=df_loc,aes(x=as.factor(need_based_aid), y=percent,fill=school_loc)) + geom_col( position="dodge") +
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(labels = c("0" = "Not Eligible", "1" = "Eligible")) +
    labs (
      x= "Eligibility of Need Based Aid",
      y = "Percentage of Students",
      fill = "Attended School Category"
      )+ geom_text(aes(label = round(percent*100,2)), vjust = -.5, position = position_dodge(width=0.9))
```

From this graph, we can see that based on the eligibility of need based aid, there is not really a relationship bewteen just the eligibility of need based aid and going to a large public school or a NW5 school. 

#### Demographics


This graph is mostly going to look at the first-generation status of students

```{r}
df_loc = df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        )) %>% 
    drop_na(first_gen_status, school_loc) %>% 
    group_by(first_gen_status, school_loc) %>% 
    summarise(subtotal = n(), .groups = "drop_last") %>% 
    mutate(total_by_var1 = sum(subtotal),
          percent = subtotal/total_by_var1) %>%
    mutate(school_loc = fct_reorder(school_loc, percent, .fun = identity, .desc=TRUE)) %>%
          #filtering out the "other schools"
          filter(school_loc == 'Large Public' | school_loc == 'NW 5') 
    print(df_loc)
    ggplot(data=df_loc,aes(x=as.factor(first_gen_status), y=percent,fill=school_loc)) + geom_col( position="dodge") +
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(labels = c("0" = "Not First Generation", "1" = "First Generation")) +
    labs (
      x= "First Generation Status",
      y = "Percentage of Students",
      fill = "Attended School Category"
      )+ geom_text(aes(label = round(percent*100,2)), vjust = -.5, position = position_dodge(width=0.9))
```

```{r}
df_loc = df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        )) %>% 
    drop_na(first_gen_status, school_loc) %>% 
    group_by(first_gen_status, school_loc) %>% 
    summarise(subtotal = n(), .groups = "drop_last") %>% 
    mutate(total_by_var1 = sum(subtotal),
          percent = subtotal/total_by_var1) %>%
    mutate(school_loc = fct_reorder(school_loc, percent, .fun = identity, .desc=TRUE)) 
          #filtering out the "other schools"
          #filter(school_loc == 'Large Public' | school_loc == 'NW 5') 
    print(df_loc)
    ggplot(data=df_loc,aes(x=as.factor(first_gen_status), y=percent,fill=school_loc)) + geom_col( position="dodge") +
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(labels = c("0" = "Not First Generation", "1" = "First Generation")) +
    labs (
      x= "First Generation Status",
      y = "Percentage of Students",
      fill = "Attended School Category"
      )+ geom_text(aes(label = round(percent*100,2)), vjust = -.5, position = position_dodge(width=0.9))
```

Learned from first-gen graphs:


Race/Ethnicity
column: ipeds_classification

```{r}
unique(df['ipeds_classification'])
```

```{r}
df_loc = df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        )) %>% 
    drop_na(ipeds_classification, school_loc) %>% 
    group_by(ipeds_classification, school_loc) %>% 
    summarise(subtotal = n(), .groups = "drop_last") %>% 
    mutate(total_by_var1 = sum(subtotal),
          percent = subtotal/total_by_var1) %>%
    mutate(school_loc = fct_reorder(school_loc, percent, .fun = identity, .desc=TRUE)) 

    print(df_loc)
    
    ggplot(data=df_loc,aes(x=ipeds_classification, y=percent,fill=school_loc)) + geom_col( position="dodge") +
    scale_y_continuous(labels = scales::percent) +
    labs (
      x= "Ethnicity of Student",
      y = "Percentage of Students",
      fill = "Attended School Category"
      )+ geom_text(aes(label = round(percent*100,2)), vjust = -.5, position = position_dodge(width=0.9)) +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

```{r}
df_loc = df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        )) %>% 
    drop_na(ipeds_classification, school_loc) %>% 
    group_by(ipeds_classification, school_loc) %>% 
    summarise(subtotal = n(), .groups = "drop_last") %>% 
    mutate(total_by_var1 = sum(subtotal),
          percent = subtotal/total_by_var1) %>%
    mutate(school_loc = fct_reorder(school_loc, percent, .fun = identity, .desc=TRUE)) %>%
          #filtering out the "other schools"
          filter(school_loc == 'Large Public' | school_loc == 'NW 5') 
    print(df_loc)
    ggplot(data=df_loc,aes(x=ipeds_classification, y=percent,fill=school_loc)) + geom_col( position="dodge") +
    scale_y_continuous(labels = scales::percent) +
    labs (
      x= "Ethnicity of Student",
      y = "Percentage of Students",
      fill = "Attended School Category"
      )+ geom_text(aes(label = round(percent*100,2)), vjust = -.5, position = position_dodge(width=0.9)) +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

#### Academics
Mostly interested in GPA, test scores, academic rigor, writing rating. Do more academically successful students tend to go private or public?

Academic columns:
app_weighted_gpa
app_unweighted_gpa
app_transfer_gpa
co_curricular_rating (1-6) 6 most demanding
curriculum_rating (1-6) 6 most demanding
writing_rating (1-4) 4 outstanding
converted_satr_superscore
person_calculated_merit_rank (num scale)
edi_access_check 1/0 Indicates whether the applicant meets EDI access criteria


```{r}
#this graph will look at the combined GPA 
df_loc = df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        ), 
        combo_gpa = case_when(
          !is.na(app_transfer_gpa) & is.na(app_unweighted_gpa) ~ app_transfer_gpa +1 ,
          is.na(app_transfer_gpa) & !is.na(app_weighted_gpa) ~ app_weighted_gpa,
          !is.na(app_weighted_gpa) & !is.na(app_transfer_gpa) ~ (((app_transfer_gpa+1)+ app_weighted_gpa) /2)
        )) %>% 
    mutate(combo_gpa = cut(combo_gpa, breaks = c(2,2.5,2.75,3,3.25,3.5,3.75,4, 4.25,4.5,4.75,5), right = FALSE)) %>%
        #mutate(unweight_gpa = cut(unweight_gpa, breaks=8)) %>%
    drop_na(combo_gpa, school_loc) %>% 
    group_by(combo_gpa, school_loc) %>% 
    summarise(subtotal = n(), .groups = "drop_last") %>% 
    mutate(total_by_var1 = sum(subtotal),
          percent = subtotal/total_by_var1) %>%
    mutate(school_loc = fct_reorder(school_loc, percent, .fun = identity, .desc=TRUE))
          #filtering out the "other schools"
          #filter(school_loc == 'Large Public' | school_loc == 'NW 5') 
    print(df_loc)
    ggplot(data=df_loc,aes(x=combo_gpa, y=percent,fill=school_loc)) + geom_col( position="dodge") +
    scale_y_continuous(labels = scales::percent) +
    #scale_x_discrete(labels = c("0" = "Not First Generation", "1" = "First Generation")) +
    labs (
      x= "Applicant Unweighted GPA",
      y = "Percentage of Students",
      fill = "Attended School Category"
      )+ geom_text(aes(label = round(percent*100,2)), vjust = -.5, position = position_dodge(width=0.9))
```

```{r}
ggplot(data=df_loc,aes(x=combo_gpa, y=percent,fill=school_loc)) + geom_col( position="dodge") +
    scale_y_continuous(labels = scales::percent) +
    #scale_x_discrete(labels = c("0" = "Not First Generation", "1" = "First Generation")) +
    labs (
      x= "Applicant Unweighted GPA",
      y = "Percentage of Students",
      fill = "Attended School Category"
      )+ geom_text(aes(label = round(percent*100,2)), vjust = -.5, position = position_dodge(width=0.9))
```

```{r}
df_loc =df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        ), 
        unweight_gpa = case_when(
          !is.na(app_unweighted_gpa) ~ app_unweighted_gpa,
          is.na(app_unweighted_gpa) & !is.na(app_transfer_gpa) ~ app_transfer_gpa
        )
        ) %>% 
        filter(unweight_gpa <= 4) %>%
        mutate(unweight_gpa = cut(unweight_gpa, breaks = c(2,2.5,2.75,3,3.25,3.5,3.75,4), right = FALSE)) %>%
        #mutate(unweight_gpa = cut(unweight_gpa, breaks=8)) %>%
    drop_na(unweight_gpa, school_loc) %>% 
    group_by(unweight_gpa, school_loc) %>% 
    summarise(subtotal = n(), .groups = "drop_last") %>% 
    mutate(total_by_var1 = sum(subtotal),
          percent = subtotal/total_by_var1) %>%
    mutate(school_loc = fct_reorder(school_loc, percent, .fun = identity, .desc=TRUE))
          #filtering out the "other schools"
          #filter(school_loc == 'Large Public' | school_loc == 'NW 5') 
    print(df_loc)
    ggplot(data=df_loc,aes(x=unweight_gpa, y=percent,fill=school_loc)) + geom_col( position="dodge") +
    scale_y_continuous(labels = scales::percent) +
    #scale_x_discrete(labels = c("0" = "Not First Generation", "1" = "First Generation")) +
    labs (
      x= "Applicant Unweighted GPA",
      y = "Percentage of Students",
      fill = "Attended School Category"
      )+ geom_text(aes(label = round(percent*100,2)), vjust = -.5, position = position_dodge(width=0.9))
```

```{r}
df_loc =df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
        ), 
        unweight_gpa = case_when(
          !is.na(app_unweighted_gpa) ~ app_unweighted_gpa,
          is.na(app_unweighted_gpa) & !is.na(app_transfer_gpa) ~ app_transfer_gpa
        )
        ) %>% 
        filter(unweight_gpa <= 4) %>%
        mutate(unweight_gpa = cut(unweight_gpa, breaks = c(2,2.5,2.75,3,3.25,3.5,3.75,4), right = FALSE)) %>%
        #mutate(unweight_gpa = cut(unweight_gpa, breaks=8)) %>%
    drop_na(unweight_gpa, school_loc) %>% 
    group_by(unweight_gpa, school_loc) %>% 
    summarise(subtotal = n(), .groups = "drop_last") %>% 
    mutate(total_by_var1 = sum(subtotal),
          percent = subtotal/total_by_var1) %>%
    mutate(school_loc = fct_reorder(school_loc, percent, .fun = identity, .desc=TRUE)) %>%
          #filtering out the "other schools"
          filter(school_loc == 'Large Public' | school_loc == 'NW 5') 
    print(df_loc)
    ggplot(data=df_loc,aes(x=unweight_gpa, y=percent,fill=school_loc)) + geom_col( position="dodge") +
    scale_y_continuous(labels = scales::percent) +
    #scale_x_discrete(labels = c("0" = "Not First Generation", "1" = "First Generation")) +
    labs (
      x= "Applicant Unweighted GPA",
      y = "Percentage of Students",
      fill = "Attended School Category"
      )+ geom_text(aes(label = round(percent*100,2)), vjust = -.5, position = position_dodge(width=0.9))
```
  

```{r}
max(df['app_transfer_gpa'], na.rm=TRUE)
max(df['app_unweighted_gpa'], na.rm=TRUE)
sum(df$app_unweighted_gpa >4, na.rm = TRUE)
```


```{python}
df4 = df[['app_unweighted_gpa', 'curriculum_rating', 'writing_rating', 'is_nw5', 'is_large_public', 'dest_other']]

```

```{r}
library(ggplot2)
library(dplyr)
library(scales)

# 1. Summarize to get counts, then calculate percentage
plot_data = df %>%
  count(writing_rating) %>%
  mutate(pct = n / sum(n))

df_percentage <- df %>%
  group_by(writing_rating) %>%
  summarise(pct_nw5 = mean(is_nw5 == 1))
# 2. Create the plot
ggplot(df_percentage, aes(x = writing_rating, y = pct_nw5)) +
  geom_col(fill = "red3") +
  geom_text(aes(label = scales::percent(pct_nw5)), vjust = -0.5) + # Adds text labels
  scale_y_continuous(labels = scales::percent) +                # Formats y-axis
  labs(x = "Category", y = "Percent of Total")

```


```{r}
df_percent <- df %>%
  group_by(writing_rating,testing_plan) %>%
  summarise(value = sum(is_nw5), total=n()) %>%
  mutate(
    total2 = sum(value),
    percent = value / total
  ) %>%
  drop_na(testing_plan)

ggplot(df_percent, aes(x = writing_rating, fill = factor(testing_plan), y=percent)) +
  geom_col(position = position_dodge()) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(
    title = "Percentage of Applicants attending Northwest 5 schools by testing plan and Writing Rating",
    subtitle = "Split by Writing Rating and Testing Plan",
    x = "Writing Rating",
    y = "Percentage of students attending a Northwest 5 school",
    fill = "Submitted Test Scores"
  )+
  scale_fill_manual(values=c("red3", "yellow"), labels=c("Submitted", "Did not submit") )
```


```{r}
#just looking at testing plan
# x axis will be where they went to school

df_percent <- df %>%
    mutate(school_loc = case_when(
        is_nw5==1 ~ "NW 5",
        is_large_public==1 ~ "Large Public",
        dest_other==1 ~ "Other"
    )) %>%
  #  group_by(testing_plan) %>%
  #summarise(value = sum(school_loc), total=n()) %>%
  #mutate(
  #  total2 = sum(value),df
  #  Percentage = value / total) %>% 
  drop_na(testing_plan) %>%
  #count(testing_plan, school_loc) %>% 
  group_by(testing_plan) %>%
  summarize( value = sum(n), total=n()) %>%
  mutate(Percentage =  value/total, total=total)
  #(n / n()), nrow=nrow(.))

ggplot(df_percent, aes(x = school_loc, y = Percentage, fill = school_loc)) +
  geom_col() +
  facet_wrap(~ testing_plan)



```